---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rasterpolygonizer

<!-- badges: start -->
[![R-CMD-check](https://github.com/jashonnew/rasterpolygonizer/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jashonnew/rasterpolygonizer/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of rasterpolygonizer is to ...

## Installation

You can install the development version of rasterpolygonizer like so:

```{r}
# Install from GitHub
# install.packages("remotes")
#remotes::install_github("jashonnew/rasterpolygonizer")
```

This package provides three main functions:

1. fill_ground_raster() – fills ground values in a raster.

2. extract_building_edges_to_polygons() – finds edge patches in a raster.

3. clean_building_polygons() – cleans and filters building polygons.

For a full example workflow using the included sample raster, see the package vignette:
```{r}
#browseVignettes("rasterpolygonizer")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(rasterpolygonizer)
library(terra)
library(sf)
library(ggplot2)

r_path <- system.file("extdata", "sample_raster.tif", package = "rasterpolygonizer")

load_extdata <- function(filename, pkg = "rasterpolygonizer") {
  path <- system.file("extdata", filename, package = pkg)
  if (path == "") stop(sprintf("File '%s' not found in inst/extdata", filename))
  path
}


r <- terra::rast(r_path)
b <- sf::st_read(load_extdata("sample_buildings.gpkg"), quiet = TRUE)


r_df_raw <- as.data.frame(r, xy = TRUE, na.rm = TRUE)
names(r_df_raw)[3] <- "value"

ggplot(r_df_raw) +
  geom_raster(aes(x = x, y = y, fill = value)) +
  coord_equal() +
  scale_fill_viridis_c(na.value = "transparent") +
  theme_minimal() +
  labs(title = "Input raster")

r_filled <- fill_ground_raster(r)

r_df_filled <- as.data.frame(r_filled, xy = TRUE, na.rm = TRUE)
names(r_df_filled)[3] <- "value"

ggplot(r_df_filled) +
  geom_raster(aes(x = x, y = y, fill = value)) +
  coord_equal() +
  scale_fill_viridis_c(na.value = "transparent") +
  theme_minimal() +
  labs(title = "Filled ground raster")

edges <- extract_building_edges_to_polygons(r_filled, thr_prob = .75)

# Use raster extent as mask for this example
mask_poly <- terra::as.polygons(r)
mask_sf <- sf::st_as_sf(mask_poly)

buildings_sf <- clean_building_polygons(
  closed_edges = edges$closed_edges,
  shrink_dist = -0.5,
  simplify_tol = 2.5,
  min_area = 19.99,
  max_area = 2000
)

# Print summary



r_df <- as.data.frame(r_filled, xy = TRUE, na.rm = TRUE)
names(r_df)[3] <- "value"
ggplot() +
  geom_raster(
    data = r_df,
    aes(x = x, y = y, fill = value)
  ) +
  geom_sf(
    data = buildings_sf,
    fill = NA,
    color = "red",
    linewidth = 0.6
  ) +
  coord_sf() +
  scale_fill_viridis_c(na.value = "transparent") +
  theme_minimal()

```

```{r message=FALSE, warning=FALSE}

terra::crs(r) <- "EPSG:32606"
r_crs <- terra::crs(r)

buildings_sf <- sf::st_set_crs(buildings_sf, r_crs)
b <- sf::st_set_crs(b, r_crs)

filtered <- filter_by_ground_truth(
  buildings_sf,
  b,
  threshold = 0.5
)

r_df <- as.data.frame(r_filled, xy = TRUE, na.rm = TRUE)
names(r_df)[3] <- "value"
ggplot() +
  geom_raster(
    data = r_df,
    aes(x = x, y = y, fill = value)
  ) +
  geom_sf(
    data = filtered,
    fill = NA,
    color = "red",
    linewidth = 0.6
  ) +
  coord_sf() +
  scale_fill_viridis_c(na.value = "transparent") +
  theme_minimal()

```


For detailed explanation and additional parameters, see the vignette.

The package includes a small example raster: `inst/extdata/sample_raster.tif`
This is used for testing and demonstrating workflows.
